---
title: "Posterior Prediction"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"
---

class: title-slide

```{r child = "../setup.Rmd"}
```

```{r echo=FALSE, message=FALSE}
library(bayesrules)
library(tidyverse)
library(janitor)
library(ggforce)
library(gridExtra)
```

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
Examples from [bayesrulesbook.com](https://bayesrulesbook.com)

]

---

class: middle

### Posterior prediction of new outcomes

Suppose we get our hands on 20 more artworks.
What number would you _predict_ are done by artists that are Gen X or younger?

--

- __sampling variability__ 

- __posterior variability__ in $\pi$.\index{sampling variability} 

---

class: middle

First, let $Y' = y'$ be the (yet unknown) number of the 20 new artworks that are done by Gen X or younger artists.
Thus conditioned on $\pi$, the randomness in $Y'$ can be modeled by $Y'|\pi \sim \text{Bin}(20,\pi)$ with pdf

$$f(y'|\pi) = P(Y' = y' | \pi) = \binom{20}{ y'} \pi^{y'}(1-\pi)^{20 - y'}$$
--

class: middle

$$f(y'|\pi) f(\pi|(y=14)) \; ,$$
---

```{r art-pred-1-ch8, fig.width = 15, echo = FALSE}
binom_plot <- function(pi_val){
  pred_plot <- data.frame(x = 0:20, y = dbinom(0:20, size = 20, prob = pi_val)*dbeta(pi_val,post_a, post_b))
  ggplot(pred_plot, aes(x = x, y = y)) + 
    geom_point(size = 0.7) + 
    geom_segment(aes(x = x, xend = x, y = 0, yend = y), size = 0.1) + 
    lims(y = c(0,3)) + 
    labs(x = "y'", title = substitute(paste(pi, " = ", nn), list(nn = pi_val)), y = expression(paste("f(y'|", pi, ")", "f(", pi, "|(y = 14))")))
}


g1 <- binom_plot(pi_val = lower_95) 
g2 <- binom_plot(pi_val = post_mode)
g3 <- binom_plot(pi_val = upper_95)

grid.arrange(g1, g2, g3, ncol=3)

```

---
class: middle

$f(y'|(y=14)) = \int_0^1 f(y'|\pi) f(\pi|(y=14)) d\pi$

--

$f(y'|(y=14)) = {20 \choose y'} \frac{\Gamma(110)}{\Gamma(18)\Gamma(92)}\frac{\Gamma(18+y')\Gamma(112-y')} {\Gamma(130)}  \text{ for } y' \in \{0,1,\ldots,20\}$

--

$f((y'=3)|(y=14)) = {20 \choose 3}\frac{\Gamma(110)}{\Gamma(18)\Gamma(92)}\frac{\Gamma(18+3)\Gamma(112-3)} {\Gamma(130)} = 0.2217$

---

```{r art-pred-2-ch8, echo = FALSE}
pred_pdf <- Vectorize(function(x){choose(20,x)*gamma(110)/gamma(18)/gamma(92)*gamma(18+x)*gamma(112-x) /gamma(130)})

# Confirm it's valid
#sum(pred_pdf(x = 0:20))

plot_pred <- data.frame(x = 0:20, y = pred_pdf(0:20))
ggplot(plot_pred, aes(x = x, y = y)) + 
  geom_point(size = 0.7) + 
  geom_segment(aes(x = x, xend = x, y = 0, yend = y), size = 0.25) + 
  labs(x = "y'", y = "f(y' | (y = 14))")


```

---

Let $Y'$ denote a new outcome of variable $Y$.  Further, let pdf $f(y'|\pi)$ denote the dependence of $Y'$ on $\pi$ and posterior pdf $f(\pi|y)$ denote the posterior plausibility of $\pi$ given the original data $Y = y$.  Then the posterior predictive model for $Y'$ has pdf

$$f(y'|y) = \int f(y'|\pi) f(\pi|y) d\pi \; .$$

In words, the overall chance of observing $Y' = y'$ weights the chance of observing this outcome under _any_ possible $\pi$ ( $f(y'|\pi)$ ) by the posterior plausibility of $\pi$ ( $f(\pi|y)$ ).

---

```{r echo=FALSE, cache=TRUE}
library(rstan)
# STEP 1: DEFINE the model
art_model <- "
  data {
    int<lower=0, upper=100> Y;
  }
  parameters {
    real<lower=0, upper=1> pi;
  }
  model {
    Y ~ binomial(100, pi);
    pi ~ beta(4, 6);
  }
"

# STEP 2: SIMULATE the posterior
art_sim <- stan(
  model_code = art_model, data = list(Y = 14), 
  chains = 4, iter = 5000*2, seed = 84735, refresh=FALSE)

art_chains_df <- as.data.frame(art_sim, 
  pars ="lp__", include = FALSE)
```

```{r}
head(art_chains_df, 3)
```

---

```{r}
set.seed(84735)
# Predict a value of Y' for each pi value in the chain
art_chains_df <- art_chains_df %>% 
  mutate(y_predict = rbinom(length(pi), size = 20, prob = pi))
```

--

```{r}
head(art_chains_df, 3)
```

---

```{r fig.height=5}
ggplot(art_chains_df, aes(x = y_predict)) + 
  stat_count()
```

