---
title: "Markov Chains"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"
---

class: title-slide

```{r child = "../setup.Rmd"}
```

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(rstan)
library(bayesplot)
```

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
Examples from [bayesrulesbook.com](https://bayesrulesbook.com)

]

---

class: middle

## MCMC (Markov Chain Monte Carlo) samples

- are *not* taken directly from the posterior pdf.
- are *not* independent
- In a given Markov chain, $f\left(\theta^{(i + 1)} \; | \; \theta^{(1)}, \theta^{(2)}, \ldots, \theta^{(i)}, y\right) = f\left(\theta^{(i + 1)} \; | \; \theta^{(i)}, y\right)$

---

## Step 1: Definte the Model

```{r}
bb_model <- "
  data {
    int<lower=0, upper=10> Y;
  }
  parameters {
    real<lower=0, upper=1> pi;
  }
  model {
    Y ~ binomial(10, pi);
    pi ~ beta(2, 2);
  }
"
```

---

## Step 2: Simulate the posterior

```{r cache=TRUE}
bb_sim <- stan(
  model_code = bb_model, data = list(Y = 9), 
  chains = 4, iter = 5000*2, seed = 84735)
```

---

```{r}
as.array(bb_sim, pars = "pi") %>% 
  head(4)
```

---

```{r}
as.array(bb_sim, pars = "pi") %>% 
  tail(4)
```


---

```{r}
as.array(bb_sim, pars = "pi") %>% 
  as.data.frame() %>% 
  mutate(iteration = 1:nrow(.))
```

---

```{r echo=FALSE}
as.array(bb_sim, pars = "pi") %>% 
  as.data.frame() %>% 
  mutate(iteration = 1:nrow(.)) %>% 
  slice(1:20) %>% 
  ggplot(aes(x = iteration, y = `chain:1.pi`)) +
  geom_line() +
  ylab("pi")
```

---

```{r}
bayesplot::mcmc_trace(bb_sim, pars = "pi", size = 0.5)
```

---

#### Histogram of the Markov chain values

```{r message=FALSE}
mcmc_hist(bb_sim, pars = "pi")
```

---

#### Density plot of the Markov chain values

```{r}
mcmc_dens(bb_sim, pars = "pi")
```

---

#### Density plot of the Markov chain values

```{r fig.height=5}
mcmc_dens(bb_sim, pars = "pi") + 
  stat_function(fun = dbeta, args = list(11, 3), color = "red")
```

---

#### Density plot of the individual chains

```{r}
mcmc_dens_overlay(bb_sim, pars = "pi")
```

---

### Effective Sample Size

```{r}
# Summarize the combined values of the 4 long chains
summary(bb_sim, pars = c("pi"))$summary
```

$N_{eff}$ is the number of independent sample values it would it take to produce an equivalently accurate posterior approximation.

---

#### Density plot of the individual chains

```{r echo=FALSE}
mcmc_dens_overlay(bb_sim, pars = "pi")
```

---

## R-hat

$$\text{R-hat} \approx \sqrt{\frac{\text{Var}_\text{combined}}{\text{Var}_\text{within}}} \; .$$
where $\text{Var}_\text{combined}$ is the combined variability in $\theta$ across all four chains and $\text{Var}_\text{within}$ is the typical variability within any individual chain.

--

$\text{R-hat} \approx 1$, reflects stability across the parallel chains.

--

$\text{R-hat} > 1$ indicates instability



