---
title: "Grid Approximation"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"
---

class: title-slide

```{r child = "../setup.Rmd"}
```

```{r echo=FALSE, message=FALSE}
library(bayesrules)
library(tidyverse)
library(janitor)
library(ggforce)
library(gridExtra)
```

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
Examples from [bayesrulesbook.com](https://bayesrulesbook.com)

]

---

class: middle

## Specifying the Posterior

$Y | \pi \sim \text{Bin}(10, \pi)$  
$\pi \sim \text{Beta}(2,2)$  

--

Suppose $Y = 9$

--

Then $\pi|(Y = 9) \sim \text{Beta}(11,3)$

---

## Step 1: Define a grid of 6 pi values

```{r}
pi_grid   <- seq(from = 0, to = 1, length = 6)
pi_grid
grid_data <- data.frame(pi_grid)
```

---

### Step 2: Evaluate the prior and likelihood at each pi

```{r}
grid_data <- grid_data %>% 
  mutate(prior = dbeta(pi_grid, 2, 2))

```

--

.pull-left[
```{r}
grid_data
```

]

--

.pull-right[

```{r echo = FALSE, fig.height=3}
ggplot(grid_data, 
       aes(x = pi_grid,
           y = prior)) +
  geom_point() +
    geom_segment(aes(x = pi_grid, xend = pi_grid, 
    y = 0, yend = prior))
```

]

---

### Step 2: Evaluate the prior and likelihood at each pi

```{r}
grid_data <- grid_data %>% 
  mutate(prior = dbeta(pi_grid, 2, 2)) %>% 
  mutate(likelihood = dbinom(9, 10, pi_grid))

```

--

.pull-left[
```{r}
grid_data
```

]

--

.pull-right[

```{r echo = FALSE, fig.height=3}
ggplot(grid_data, 
       aes(x = pi_grid,
           y = likelihood)) +
  geom_point() +
    geom_segment(aes(x = pi_grid, xend = pi_grid, 
    y = 0, yend = likelihood))
```
]

---

## Step 3: Approximate the posterior

```{r}
grid_data <- grid_data %>% 
  mutate(unnormalized = likelihood * prior) 

grid_data
```


---

## Step 3: Approximate the posterior

```{r}
grid_data <- grid_data %>% 
  mutate(unnormalized = likelihood * prior) %>% 
  mutate(posterior = unnormalized / sum(unnormalized))

grid_data
```

---

## Step 3: Approximate the posterior

```{r}
grid_data <- grid_data %>% 
  mutate(unnormalized = likelihood * prior) %>% 
  mutate(posterior = unnormalized / sum(unnormalized))
```

Confirm that the posterior approximation sums to 1


```{r}
grid_data %>% 
  summarize(sum(unnormalized), sum(posterior))
```

---

## The Grid Approximated Posterior

.pull-left[
```{r eval = FALSE}
ggplot(grid_data, 
       aes(x = pi_grid, y = posterior)) + 
  geom_point() + 
  geom_segment(aes(x = pi_grid, 
                   xend = pi_grid, 
                   y = 0, 
                   yend = posterior))
```

]

.pull-right[
```{r echo = FALSE, fig.height=5}
ggplot(grid_data, aes(x = pi_grid, y = posterior)) + 
  geom_point() + 
  geom_segment(aes(x = pi_grid, xend = pi_grid, 
    y = 0, yend = posterior))
```

]

---

## Step 4: Sample from the Discretized Posterior

```{r}
set.seed(84735)
post_sample <- sample_n(grid_data, size = 10000, 
  weight = posterior, replace = TRUE)
```

--

```{r}
post_sample %>% 
  tabyl(pi_grid) %>% 
  adorn_totals("row")
```

---

## Grid Simulation with the Posterior pdf

.pull-left[
```{r message = FALSE, eval = FALSE}
ggplot(post_sample, aes(x = pi_grid)) + 
  geom_histogram(aes(y = ..density..), 
                 color = "white") + 
  stat_function(fun = dbeta, 
                args = list(11, 3)) + 
  lims(x = c(0, 1))
```
]

.pull-right[
```{r message = FALSE, echo = FALSE, warning=FALSE, fig.height=5}
ggplot(post_sample, aes(x = pi_grid)) + 
  geom_histogram(aes(y = ..density..), color = "white") + 
  stat_function(fun = dbeta, args = list(11, 3)) + 
  lims(x = c(0, 1))
```
]


---

Let's increase the possible $\pi$ values from 6 to 101.

```{r}
pi_grid   <- seq(from = 0, to = 1, length = 101)
```


```{r echo = FALSE, fig.height=5}
# Step 1: Define a grid of 101 pi values
pi_grid    <- seq(from = 0, to = 1, length = 101)
grid_data  <- data.frame(pi_grid)

# Step 2: Evaluate the prior & likelihood at each pi
grid_data <- grid_data %>% 
  mutate(prior = dbeta(pi_grid, 2, 2)) %>% 
  mutate(likelihood = dbinom(9, 10, pi_grid))

# Step 3: Approximate the posterior
grid_data <- grid_data %>% 
  mutate(unnormalized = likelihood * prior) %>% 
  mutate(posterior = unnormalized / sum(unnormalized))
```

---

## Grid simulation with posterior pdf

```{r echo=FALSE}
# Set the seed
set.seed(84735)

# Step 4: sample from the discretized posterior
post_sample <- sample_n(grid_data, size = 10000, 
  weight = posterior, replace = TRUE)
```


.pull-left[
```{r message = FALSE, eval = FALSE}
ggplot(post_sample, aes(x = pi_grid)) + 
  geom_histogram(aes(y = ..density..), 
                 color = "white") + 
  stat_function(fun = dbeta, 
                args = list(11, 3)) + 
  lims(x = c(0, 1))
```
]

.pull-right[
```{r message = FALSE, echo = FALSE, warning=FALSE, fig.height=5}
ggplot(post_sample, aes(x = pi_grid)) + 
  geom_histogram(aes(y = ..density..), 
                 color = "white") + 
  stat_function(fun = dbeta, args = list(11, 3)) + 
  lims(x = c(0, 1))
```
]


---

## Grid approximation


```{r echo=FALSE, fig.height=3}
# Alt Text: Three images of a rainbow. The first image shows only a few "slices" of the rainbow, the second image shows even more slices for a fuller image of the rainbow, and the third shows the complete rainbow.

# full rainbow
g <- ggplot() + 
  geom_circle(aes(x0 = 0, y0 = 0, r = 2), fill = "red", color = "red") + 
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.8), fill = "orange", color = "orange") + 
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.6), fill = "yellow", color = "yellow") + 
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.4), fill = "green", color = "green") + 
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.2), fill = "blue", color = "blue") + 
  geom_circle(aes(x0 = 0, y0 = 0, r = 1), fill = "purple", color = "purple") + 
  geom_circle(aes(x0 = 0, y0 = 0, r = 0.8), fill = "white", color = "white") + 
  coord_cartesian(ylim = c(0,2), xlim = c(-2,2)) + 
  theme_void()


# define rectangles
n   <- 21
x1  <- seq(-2, 2, length = n)
x2  <- c(x1[-length(x1)] + 0.1*diff(x1), max(x1))
y1  <- rep(-0.5, length(x2))
y2  <- rep(2.1, length(x2))

rect_df <- data.frame(x1, x2, y1, y2)
g3 <- g + geom_rect(data = rect_df, aes(xmin = x1, xmax = x2, ymin = y1, ymax = y2), color = "white", fill = "white")


x1new <- x1[-c(2*(1:(0.5*(length(x1)+1))))]
x2new <- x2[-1]
x2new <- c(x2new[-c(2*(1:(0.5*(length(x2new)+1))))], max(x1new))
y1new  <- rep(-0.5, length(x2new))
y2new  <- rep(2.1, length(x2new))
rect_df <- data.frame(x1=x1new, x2=x2new, y1=y1new, y2=y2new)
g2 <- g + geom_rect(data = rect_df, aes(xmin = x1, xmax = x2, ymin = y1, ymax = y2), color = "white", fill = "white")


x1new <- x1new[-c(2*(1:(0.5*(length(x1new)+1))))]
x2new <- x2new[-1]
x2new <- c(x2new[-c(2*(1:(0.5*(length(x2new)+1))))], max(x1new))
y1new  <- rep(-0.5, length(x2new))
y2new  <- rep(2.1, length(x2new))
rect_df <- data.frame(x1=x1new, x2=x2new, y1=y1new, y2=y2new)
g1 <- g + geom_rect(data = rect_df, aes(xmin = x1, xmax = x2, ymin = y1, ymax = y2), color = "white", fill = "white")

#grid.arrange(g1,g2,g3,g,ncol=2)
grid.arrange(g1, g3, g, ncol=3)
```

---

### Grid approximation

Grid approximation produces a sample of $N$ __independent__ $\theta$ values, $\left\lbrace \theta^{(1)}, \theta^{(2)}, \ldots, \theta^{(N)} \right\rbrace$, from a __discretized approximation__ of posterior pdf $f(\theta|y)$.  This algorithm evolves in four steps:

1. Define a discrete grid of possible $\theta$ values.
2. Evaluate the prior pdf $f(\theta)$ and likelihood function $L(\theta|y)$ at each $\theta$ grid value.
3. Obtain a discrete approximation of posterior pdf $f(\theta |y)$ by: (a) calculating the product $f(\theta)L(\theta|y)$ at each $\theta$ grid value; and then (b) *normalizing* the products so that they sum to 1 across all $\theta$.
4. Randomly sample $N$ $\theta$ grid values with respect to their corresponding normalized posterior probabilities.

