---
title: "Extending the Normal Regression"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"
---

class: title-slide

```{r child = "../setup.Rmd"}
```

```{r echo=FALSE, message=FALSE}
library(bayesrules)
library(tidyverse)
library(rstanarm)
library(bayesplot)
library(janitor)
library(rstan)
```

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
Examples from [bayesrulesbook.com](https://bayesrulesbook.com)

]

---

```{r}
weather_WU <- weather_australia %>% 
  filter(location %in% c("Wollongong", "Uluru")) %>%
  mutate(location = droplevels(as.factor(location))) %>% 
  select(location, windspeed9am, humidity9am, 
    pressure9am, temp9am, temp3pm)
```

--

```{r}
glimpse(weather_WU)
```

---

```{r fig.height=6}
ggplot(weather_WU, aes(x = temp9am, y = temp3pm)) +
  geom_point()
```

---

class: middle

$\text{likelihood model:} \; \; \; Y_i | \beta_0, \beta_1, \sigma \;\;\;\stackrel{ind}{\sim} N\left(\mu_i, \sigma^2\right)\text{ with } \mu_i = \beta_0 + \beta_1X_{i1}$

$\text{prior models:}$ 

$\beta_0\sim N(\ldots, \ldots )$  
$\beta_1\sim N(\ldots, \ldots )$  
$\sigma \sim \text{Exp}(\ldots)$


---

class: middle

```{r eval=FALSE}
weather_model_1 <- stan_glm(temp3pm ~ temp9am, 
                            data = weather_WU,
                            family = gaussian,
                            chains = 4, 
                            iter = 5000*2, 
                            seed = 84735)
```

```{r echo=FALSE, cache=TRUE}
weather_model_1 <- stan_glm(
temp3pm ~ temp9am, data = weather_WU,
family = gaussian,
chains = 4, iter = 5000*2, seed = 84735, refresh=FALSE)

```

---

```{r fig.width=15}
mcmc_dens_overlay(weather_model_1)
```

---

```{r}
posterior_interval(weather_model_1, prob = 0.80)
```

---

```{r}
pp_check(weather_model_1)
```

---

### Considering a categorical predictor

```{r fig.width=10, fig.height=5}
ggplot(weather_WU, aes(x = temp3pm, fill = location)) +
  geom_density(alpha = 0.5)
```

---

class: middle

$$X_{i2} = \begin{cases}
1 & \text{ Wollongong} \\
0 & \text{ otherwise (ie. Uluru)} \\
\end{cases}$$

---

class: middle

$\text{likelihood model:} \; \; \; Y_i | \beta_0, \beta_1, \sigma \;\;\;\stackrel{ind}{\sim} N\left(\mu_i, \sigma^2\right)\text{ with } \mu_i = \beta_0 + \beta_1X_{i2}$

$\text{prior models:}$ 

$\beta_0\sim N(\ldots, \ldots )$  
$\beta_1\sim N(\ldots, \ldots )$  
$\sigma \sim \text{Exp}(\ldots)$

--

For Uluru, $X_{i2} = 0$ and the trend in 3pm temperature simplifies to 

$$\beta_0 + \beta_1 \cdot 0 = \beta_0 \; .$$
For Wollongong, $X_{i2} = 1$ and the trend in 3pm temperature is

$$\beta_0 + \beta_1 \cdot 1 = \beta_0 + \beta_1 \; .$$
---

class: middle

### Simulating the Posterior

```{r eval = FALSE}
weather_model_2 <- stan_glm(temp3pm ~ location, 
                            data = weather_WU, 
                            family = gaussian, 
                            chains = 4, 
                            iter = 5000*2, 
                            seed = 84735)
```

```{r echo = FALSE, cache=TRUE}
weather_model_2 <- stan_glm(
  temp3pm ~ location, data = weather_WU, 
  family = gaussian, 
  chains = 4, iter = 5000*2, seed = 84735, refresh = FALSE)
```

---

```{r fig.width=15}
mcmc_dens_overlay(weather_model_2)
```

---

```{r}
# Posterior summary statistics
model_summary <- summary(weather_model_2)
head(as.data.frame(model_summary), -2) %>% 
  select(mean, "10%", "90%", Rhat)
```

--

```{r}
b0 <- model_summary[1,1]
b1 <- model_summary[2,1]
```

---

```{r fig.width = 10, fig.height=5}
ggplot(weather_WU, aes(x = temp3pm, fill = location)) + 
  geom_density(alpha = 0.5) + 
  geom_vline(xintercept = c(b0, b0 + b1), 
    linetype = "dashed")
```

---

### Two Predictors

```{r message=FALSE, fig.height=5}
ggplot(weather_WU, 
    aes(y = temp3pm, x = temp9am, color = location)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

---

class: middle

$\text{likelihood model:}$
$Y_i | \beta_0, \beta_1, \beta_2 \sigma \;\;\;\stackrel{ind}{\sim} N\left(\mu_i, \sigma^2\right)\text{ with } \mu_i = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2}$

$\text{prior models:}$ 

$\beta_0\sim N(m_0, s_0 )$  
$\beta_1\sim N(m_1, s_1 )$  
$\beta_2\sim N(m_2, s_2 )$  
$\sigma \sim \text{Exp}(l)$

---

class: middle 

In _Uluru_, $X_{i2} = 0$ and the trend in the relationship between 3pm and 9am temperature simplifies to

$$\beta_0 + \beta_1 X_{i1} + \beta_2 \cdot 0 = \beta_0 + \beta_1 X_{i1} \; .$$

In _Wollongong_, $X_{i2} = 1$ and the trend in the relationship between 3pm and 9am temperature simplifies to

$$\beta_0 + \beta_1 X_{i1} + \beta_2 \cdot 1 = (\beta_0 + \beta_2) + \beta_1 X_{i1} \; .$$
---

class: middle 

```{r eval = FALSE}
weather_model_3 <- stan_glm(temp3pm ~ temp9am + location, 
                            data = weather_WU, 
  
                            family = gaussian, 
                            chains = 4, 
                            iter = 5000*2, 
                            seed = 84735)
```

```{r echo = FALSE, cache=TRUE}
weather_model_3 <- stan_glm(
  temp3pm ~ temp9am + location, data = weather_WU, 
  family = gaussian, 
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0)
```

---


```{r}
weather_model_3_df <- as.data.frame(weather_model_3)
head(weather_model_3_df, 3)
```

---

```{r eval=FALSE}
first_50 <- head(weather_model_3_df, 50)

ggplot(weather_WU, aes(x = temp9am, y = temp3pm)) + 
  geom_point(size = 0.01) + 
  geom_abline(data = first_50, size = 0.1,
    aes(intercept = `(Intercept)`, slope = temp9am)) + 
  geom_abline(data = first_50, size = 0.1,
    aes(intercept = `(Intercept)` + locationWollongong, 
    slope = temp9am), color = "blue")
```

---

```{r echo=FALSE}
first_50 <- head(weather_model_3_df, 50)

ggplot(weather_WU, aes(x = temp9am, y = temp3pm)) + 
  geom_point(size = 0.01) + 
  geom_abline(data = first_50, size = 0.1,
    aes(intercept = `(Intercept)`, slope = temp9am)) + 
  geom_abline(data = first_50, size = 0.1,
    aes(intercept = `(Intercept)` + locationWollongong, 
    slope = temp9am), color = "blue")
```


```{r echo = FALSE}
stats <- posterior_interval(weather_model_3, prob = 0.80)
```

---

```{r cache=TRUE}
# Simulate a set of predictions
set.seed(84735)
temp3pm_prediction <- posterior_predict(
  weather_model_3, 
  newdata = data.frame(
    temp9am = c(10, 10), location = c("Uluru", "Wollongong")))
```

---

### Posterior Predictive Model

```{r fig.height=4}
shortcut_df <- data.frame(uluru = temp3pm_prediction[,1],
                          woll = temp3pm_prediction[,2])
ggplot(shortcut_df, aes(x = uluru)) +
  geom_density() +
  geom_density(aes(x = woll), color = "blue")
```

---

```{r message=FALSE, fig.height=4}
ggplot(weather_WU,
       aes(y = temp3pm, x = humidity9am, color = location)) + 
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE)
```

---

class: middle

$\text{likelihood model:}$
$Y_i | \beta_0, \beta_1, \beta_2, \beta_3 \sigma \;\;\;\stackrel{ind}{\sim} N\left(\mu_i, \sigma^2\right)\text{ with }$ 
$\mu_i = \beta_0 + \beta_1X_{i2} + \beta_2X_{i3} + \beta_3X_{i2}X_{i3}$

$\text{prior models:}$ 

$\beta_0\sim N(m_0, s_0 )$  
$\beta_1\sim N(m_1, s_1 )$  
$\beta_2\sim N(m_2, s_2 )$  
$\beta_3\sim N(m_3, s_3 )$  
$\sigma \sim \text{Exp}(l)$

---

In _Uluru_, $X_{2} = 0$ and the trend in the relationship between temperature and humidity simplifies to

$$\mu = \beta_0 + \beta_2 X_{3} \; .$$

In _Wollongong_, $X_{2} = 1$ and the trend in the relationship between temperature and humidity simplifies to

$$\mu = \beta_0 + \beta_1 + \beta_2 X_{3} + \beta_3 X_{3} = (\beta_0 + \beta_1) + (\beta_2 + \beta_3) X_3 \; .$$

---

class: middle

```{r eval=FALSE}
interaction_model <- stan_glm(temp3pm ~ location + humidity9am + 
                                location:humidity9am,
                              data = weather_WU, 
                              family = gaussian,
                              chains = 4, 
                              iter = 5000*2, 
                              seed = 84735)
```

```{r echo=FALSE, cache=TRUE}
interaction_model <- stan_glm(
temp3pm ~ location + humidity9am + location:humidity9am,
data = weather_WU, family = gaussian,
chains = 4, iter = 5000*2, seed = 84735, refresh=FALSE)
```

---

```{r}
model_summary <- summary(interaction_model)
head(as.data.frame(model_summary), -2) %>%
select(`10%`, `50%`, `90%`) %>%
round(3)
```


```{r echo = FALSE, cache=TRUE}
interaction_model <- stan_glm(
  temp3pm ~ location + humidity9am + location:humidity9am, 
  data = weather_WU, family = gaussian, 
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0)

model_summary <- summary(interaction_model)
post_sum <- as.data.frame(model_summary) %>% 
  select(`10%`, `50%`, `90%`) %>% 
  round(3) %>% 
  head(-2)
```


---

class: middle

$$\begin{array}{lrl}
\text{Uluru:}      & \mu & =  `r post_sum[1,2]` - `r -post_sum[3,2]` \text{ humidity9am} \\
\text{Wollongong:} & \mu & = (`r post_sum[1,2]` - `r -post_sum[2,2]`) + (`r post_sum[3,2]` + `r post_sum[4,2]`) \text{ humidity9am}\\
&& = `r post_sum[1,2] + post_sum[2,2]` + `r post_sum[3,2] + post_sum[4,2]` \text{ humidity9am}\\
\end{array}$$

---

## Do you need an interaction term?

- __Context.__ 

- __Visualizations.__ 

- __Hypothesis tests.__ 

---

## More than two predictors

$\text{likelihood model:} \; \; \; Y_i | \beta_0, \beta_1,\beta_2,...\beta_p, \sigma \;\;\;\stackrel{ind}{\sim} N\left(\mu_i, \sigma^2\right)\text{ with }$ 
$\mu_i = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + \ldots +\beta_pX_{ip}$

$\text{prior models:}$ 

$\beta_0, \beta_1,\beta_2, ...,\beta_p\sim N(\ldots, \ldots )$  
 
$\sigma \sim \text{Exp}(\ldots)$

---

class: middle

```{r eval = FALSE}
weather_model_4 <- stan_glm(temp3pm ~ ., 
                            data = weather_WU, 
                            family = gaussian, 
                            chains = 4, 
                            iter = 5000*2, 
                            seed = 84735)
```

```{r echo = FALSE, cache=TRUE}
weather_model_4 <- stan_glm(
  temp3pm ~ ., data = weather_WU, 
  family = gaussian, 
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0)
```

---

class: middle center 

## Model evaluation and comparison

<div align="center">

| Model             | Formula                        |
|-------------------|--------------------------------|
| `weather_model_1` | `temp3pm ~ temp9am`            |
| `weather_model_2` | `temp3pm ~ location`           |
| `weather_model_3` | `temp3pm ~ temp9am + location` |
| `weather_model_4` | `temp3pm ~ .`                  |

---

```{r echo = FALSE, fig.width = 10, fig.height = 7}
g1 <- pp_check(weather_model_1) + theme(legend.position = "bottom")
g2 <- pp_check(weather_model_2) + theme(legend.position = "bottom")
g3 <- pp_check(weather_model_3) + theme(legend.position = "bottom")
g4 <- pp_check(weather_model_4) + theme(legend.position = "bottom")
gridExtra::grid.arrange(g1,g2,g3,g4,ncol = 2)
```

---

```{r cache=TRUE}
set.seed(84735)
predictions_1 <- posterior_predict(weather_model_1, 
  newdata = weather_WU)
```

```{r echo = FALSE, cache=TRUE}
set.seed(84735)
predictions_2 <- posterior_predict(weather_model_2, 
  newdata = weather_WU)

set.seed(84735)
predictions_3 <- posterior_predict(weather_model_3, 
  newdata = weather_WU)

set.seed(84735)
predictions_4 <- posterior_predict(weather_model_4, 
  newdata = weather_WU)
```

```{r fig.width = 4, fig.height = 2.25, eval = FALSE}
# Posterior predictive models for weather_model_1
ppc_intervals(weather_WU$temp3pm, 
  yrep = predictions_1, 
  x = weather_WU$temp9am, 
  prob = 0.5, prob_outer = 0.95)

# Posterior predictive models for weather_model_2
ppc_violin_grouped(weather_WU$temp3pm, 
  yrep = predictions_2, 
  group = weather_WU$location,
  y_draw = "points")
```


---

```{r eval=FALSE}
prediction_summary_cv(data = weather_WU, 
                      model = weather_model_1, 
                      k = 10)
```

---

class: middle center

```{r echo = FALSE}
set.seed(84735)
ps_1 <- prediction_summary_cv(
  data = weather_WU, model = weather_model_1, k = 10)$cv

set.seed(84735)
ps_2 <- prediction_summary_cv(
  data = weather_WU, model = weather_model_2, k = 10)$cv

set.seed(84735)
ps_3 <- prediction_summary_cv(
  data = weather_WU, model = weather_model_3, k = 10)$cv

set.seed(84735)
ps_4 <- prediction_summary_cv(
  data = weather_WU, model = weather_model_4, k = 10)$cv

post_sum_df <- data.frame(rbind(round(ps_1,3), round(ps_2,3), round(ps_3,3), round(ps_4,3)))
```


```{r echo = FALSE, warning=FALSE}
test_mat <- matrix(
  c("model", "mae", "mae scaled", "within 50", "within 95", 
    "weather model 1", post_sum_df[1,],
    "weather model 2", post_sum_df[2,],
    "weather model 3", post_sum_df[3,],
    "weather model 4", post_sum_df[4,]
    ), nrow = 5, byrow = TRUE)

knitr::kable(test_mat)
```



