---
title: "The Beta-Binomial Model"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"
---

class: title-slide

```{r child = "../setup.Rmd"}
```

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
Examples from [bayesrulesbook.com](https://bayesrulesbook.com)

]


---

## Bike ownership
The transportation office at our school wants to know $\pi$ the proportion of people who own bikes on campus so that they can build bike racks accordingly. The staff at the office think that the $\pi$ is more likely to be somewhere 0.05 to 0.25. The plot below shows their prior distribution. Write out a reasonable $f(\pi)$. Calculate the prior expected value, mode, and variance. 

```{r echo = FALSE}
library(ggplot2)
library(bayesrules)
```


```{r fig.align='center', fig.height=3.5, echo = FALSE}
plot_beta(5, 35) +
    theme(text = element_text(size=20)) 
```

---

## Plotting the prior

```{r eval = FALSE}
plot_beta(5, 35) 
```

```{r fig.align='center', fig.height=4, echo = FALSE}
plot_beta(5, 35) +
    theme(text = element_text(size=20)) 
```

---

## Summarizing the prior

```{r}
summarize_beta(5, 35)
```


---

## Posterior for the Beta-Binomial Model

Let $\pi \sim \text{Beta}(\alpha, \beta)$ and $Y|n \sim \text{Bin}(n,\pi)$. 

--

$f(\pi|y) \propto \frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\pi^{\alpha-1} (1-\pi)^{\beta-1} {n \choose y}\pi^y(1-\pi)^{n-y}$

--


$f(\pi|y) \propto \pi^{(\alpha+y)-1} (1-\pi)^{(\beta+n-y)-1}$

--

$\pi|y \sim \text{Beta}(\alpha +y, \beta+n-y)$

--

$f(\pi|y) = \frac{\Gamma(\alpha+\beta+n)}{\Gamma(\alpha+y)\Gamma(\beta+n-y)} \pi^{(\alpha+y)-1} (1-\pi)^{(\beta+n-y)-1}$

---

## Conjugate prior

We say that $f(\pi)$ is a conjugate prior for $L(\pi|y)$ if the posterior, $f(\pi|y) \propto f(\pi)L(\pi|y)$, is from the same model family as the prior.  

Thus, Beta distribution is a conjugate prior for the Binomial likelihood model since the posterior also follows a Beta distribution.

---

## Bike ownership posterior

The transportation office decides to collect some data and samples 50 people on campus and asks them whether they own a bike or not. It turns out that 25 of them do! What is the posterior distribution of $\pi$ after having observed this data? 

--

$\pi|y \sim \text{Beta}(\alpha +y, \beta+n-y)$

--

$\pi|y \sim \text{Beta}(5 +25, 35+50-25)$

--

$\pi|y \sim \text{Beta}(30, 60)$


---

## Plotting the posterior

```{r eval = FALSE}
plot_beta(30, 60) 
```

```{r fig.align='center', fig.height=4, echo = FALSE}
plot_beta(30, 60) +
    theme(text = element_text(size=20)) 
```


---

## Summarizing the posterior

```{r}
summarize_beta(30,60)
```



---

## Plot summary

```{r eval = FALSE}
plot_beta(30, 60, mean = TRUE, mode = TRUE) 
```

```{r fig.align='center', fig.height=4, echo = FALSE}
plot_beta(30, 60, mean = TRUE, mode = TRUE) +
    theme(text = element_text(size=20)) 
```

---

## Bike ownership: balancing act

```{r eval = FALSE}
plot_beta_binomial(alpha = 5, beta = 35,
                   y = 25, n = 50)
```

```{r fig.align='center', fig.height=5, echo = FALSE}
plot_beta_binomial(alpha = 5, beta = 35,
                   y = 25, n = 50) +
      theme(text = element_text(size=20)) 

```



---

## Posterior Descriptives

$\pi|(Y=y) \sim \text{Beta}(\alpha+y, \beta+n-y)$

$$E(\pi | (Y=y)) = \frac{\alpha + y}{\alpha + \beta + n}$$ 
$$\text{Mode}(\pi | (Y=y))  = \frac{\alpha + y - 1}{\alpha + \beta + n - 2} $$
$$\text{Var}(\pi | (Y=y))   = \frac{(\alpha + y)(\beta + n - y)}{(\alpha + \beta + n)^2(\alpha + \beta + n + 1)}\\$$
---

## Bike ownership - descriptives of the posterior

What is the descriptive measures (expected value, mode, and variance) of the posterior distribution for the bike ownership example?

--


```{r}
summarize_beta_binomial(5, 35, y = 25, n = 50)
```

