---
title: "The Gamma-Poisson Models"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"
---

class: title-slide

```{r child = "../setup.Rmd"}
```

```{r echo=FALSE, message=FALSE}
library(bayesrules)
library(tidyverse)
```

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
Examples from [bayesrulesbook.com](https://bayesrulesbook.com)

]

---

class: middle

## Choosing a Prior

Prior distribution depends on our current information. When choosing a prior we may consider 

- computational ease
- interpretability

---

## Conjugate Prior

Let the prior model for parameter $\theta$ have pdf $f(\theta)$ and the model of data $Y$ conditioned on $\theta$ have likelihood function $L(\theta|y)$.  If the resulting posterior model with pdf $f(\theta|y) \propto f(\theta)L(\theta|y)$ is of the same model family as the prior, then we say this is a __conjugate prior__. 

--

__Examples__

The Beta-Binomial Model - Beta is a conjugate prior for the Binomial Likelihood.

--

The Gamma-Poisson Model 

--

The Normal-Normal Model

---

## Non-Conjugate Prior for Binomial Likelihood

$$f(\pi)=e-e^\pi\; \text{ for } \pi \in [0,1] $$
--

- Is this a valid pdf?


$f(\pi)$ is non-negative on the support of $\pi$.  `r fontawesome::fa("check")`


```{r echo=FALSE, fig.height=4}
pi<-seq(0, 1, by= .01)
f.pi<-exp(1)-exp(pi)
datas<-as.data.frame(cbind(pi, f.pi))
g <- ggplot(datas, aes(x = pi, y = f.pi)) + 
  lims(x = c(0,1), y = c(0,2)) + 
  geom_line() + 
  labs(x = expression(pi), y = expression(paste("f(",pi,")"))) 
g 
``` 

---

## Non-Conjugate Prior for Binomial Likelihood

$$f(\pi)=e-e^\pi\; \text{ for } \pi \in [0,1] $$
--

- Is this a valid pdf?

$\int_0^1 f(\pi) \stackrel{?}{=} 1$

--

$\int_0^1 e-e^\pi\ d\pi=(e\pi - e^\pi)|_0^1 = [e-0] -[e-1]= 1$

--

$\int_0^1 f(\pi) = 1$ `r fontawesome::fa("check")`


---

## Getting to the posterior model

Assume $Y = 10$ and  $n = 50$

$L(\pi | (y=10)) = {50 \choose 10} \pi^{10} (1-\pi)^{40} \; \; \text{ for } \pi \in [0,1] \; .$

--

$f(\pi | (y = 10)) \propto f(\pi) L(\pi | (y = 10)) = (e-e^\pi) \cdot \binom{50}{10} \pi^{10} (1-\pi)^{40}.$

--

$f(\pi | y ) \propto (e-e^\pi)  \pi^{10} (1-\pi)^{40}.$

--


$f(\pi|y=10)= \frac{(e-e^\pi)  \pi^{10} (1-\pi)^{40}}{\int_0^1(e-e^\pi)  \pi^{10} (1-\pi)^{40}d\pi}  \; \; \text{ for } \pi \in [0,1].$

We would need to integrate to calculate this posterior model, and integrate again for its mean, mode, and variance.

---

## The Question

We are interested in modeling $\lambda$ the rate of fraud risk calls received per day. We plan on collecting data on the number of fraud risk phone calls received each day.

---

__The Poisson model__

Let random variable $Y$ be the _number of independent events_ that occur in a fixed amount of time or space, where $\lambda > 0$ is the rate at which these events occur.  Then the _dependence_ of $Y$ on __parameter__ $\lambda$ can be modeled by the Poisson.  In mathematical notation:

$$Y | \lambda \sim \text{Pois}(\lambda) $$

--

The Poisson model is specified by a conditional pmf:    

\begin{equation}
f(y|\lambda) =  \frac{\lambda^y e^{-\lambda}}{y!}\;\; \text{ for } y \in \{0,1,2,\ldots\}
\end{equation}

--

A Poisson random variable $Y$ is assumed to have equal mean and variance,

\begin{equation}
E(Y|\lambda) = \text{Var}(Y|\lambda) = \lambda \;
\end{equation}


---

__Joint likelihood function__

Let $(Y_1,Y_2,\ldots,Y_n)$ be an independent sample of random variables and $\vec{y} = (y_1,y_2,\ldots,y_n)$ be the corresponding vector of observed values. 

\begin{equation}
L(\lambda | \vec{y}) = \prod_{i=1}^n L(\lambda | y_i) = L(\lambda | y_1) \cdot L(\lambda | y_2) \cdots L(\lambda | y_n) \; 
\end{equation}

--

\begin{equation}
L(\lambda | \vec{y}) = \prod_{i=1}^{n}f(y_i | \lambda) = \prod_{i=1}^{n}\frac{\lambda^{y_i}e^{-\lambda}}{y_i!} \;\; \text{ for } \; \lambda > 0 \; 
\end{equation}

---

__Joint likelihood function__


$$\begin{split}
L(\lambda | \vec{y}) 
& = \prod_{i=1}^{n}\frac{\lambda^{y_i}e^{-\lambda}}{y_i!} \\
& = \frac{\lambda^{y_1}e^{-\lambda}}{y_1!} \cdot \frac{\lambda^{y_2}e^{-\lambda}}{y_2!} \cdots \frac{\lambda^{y_n}e^{-\lambda}}{y_n!} \\
& =\frac{\lambda^{\sum y_i}e^{-n\lambda}}{\prod_{i=1}^n y_i!} \\
\end{split}$$

---

## Poisson Likelihood

We collect four days of data and receive 6, 2, 2, 1 spam calls each day. Write out the likelihood model.

$$L(\lambda | \vec{y}) =\frac{\lambda^{\sum y_i}e^{-n\lambda}}{\prod_{i=1}^n y_i!}$$

--

$$L(\lambda | \vec{y}) = \frac{\lambda^{6 +2+2+1}e^{-4\lambda}}{6!\times2!\times2!\times1!} \propto \lambda^{11}e^{-4\lambda} \; .$$ 
---

```{r fig.height=5}
plot_poisson_likelihood(y = c(6, 2, 2, 1), 
                        lambda_upper_bound = 10)
```

---

## Gamma Prior


Let $\lambda$ be a random variable which can take any value between 0 and $\infty$, ie. $\lambda \in [0,\infty)$.  Then the variability in $\lambda$ might be well modeled by a Gamma model with __shape parameter__ $s > 0$ and __rate parameter__ 
$r > 0$: 

$$\lambda \sim \text{Gamma}(s, r)$$

--

The Gamma model is specified by continuous pdf

\begin{equation}
f(\lambda) = \frac{r^s}{\Gamma(s)} \lambda^{s-1} e^{-r\lambda} \;\; \text{ for } \lambda > 0 
\end{equation}

where constant $\Gamma(s) = \int_0^\infty z^{s - 1} e^{-z}dz$. When $s$ is a positive integer, $s \in \{1,2,3,\ldots\}$, this constant simplifies to $\Gamma(s) = (s - 1)!$. 

---

## The Exponential model

$$\lambda \sim \text{Exp}(r)$$

is a special case of the Gamma with shape parameter $s = 1$, $\text{Gamma}(1,r)$.

---

```{r echo=FALSE, out.width="55%"}

# Set up gamma data
alpha <- rep(c(1,2,4), 2)
beta  <- rep(c(1,2), each = 3)
gammas <- data.frame(setting = factor(rep(1:6, each = 500)), 
  x = rep(seq(0, 7.5, length = 500), 6),
  alpha = rep(alpha, each = 500), 
  beta = rep(beta, each = 500))
gammas <- gammas %>% 
    mutate(y = dgamma(x, shape = alpha, rate = beta))
levels(gammas$setting) <- paste0("Gamma(",alpha,",",beta,")")
trend_data <- data.frame(alpha, beta, 
  means = (alpha / (beta)), modes = ((alpha - 1) / (beta))) %>% 
  mutate(Parameter = paste0("Gamma(",alpha,",",beta,")")) %>% 
  mutate(setting = Parameter) %>% 
  mutate(means_d = dgamma(means, alpha, beta), modes_d = dgamma(modes, alpha, beta))
trend_data$setting <- factor(trend_data$setting, levels = c("Gamma(1,1)","Gamma(2,1)","Gamma(4,1)","Gamma(1,2)","Gamma(2,2)","Gamma(4,2)"))
  
g <- ggplot(gammas, aes(x = x, y = y)) + 
  lims(y = c(0,2)) + 
  geom_line() + 
  facet_wrap(~ setting) + 
  labs(x = expression(lambda), y = expression(paste("f(",lambda,")"))) 
g + 
  geom_segment(data = trend_data, aes(x = means, xend = means, y = 0, yend = means_d), color = "blue") + 
  geom_segment(data = trend_data, aes(x = modes, xend = modes, y = 0, yend = modes_d), linetype = "dashed", color = "blue")
``` 
---

## Trend and Variability

\begin{split}
E(\lambda) & = \frac{s}{r} \\
\text{Mode}(\lambda) & = \frac{s - 1}{r} \;\; \text{ for } s \ge 1 \\
\text{Var}(\lambda) & = \frac{s}{r^2} \\
\end{split}


---

## Gamma Model

What is $f(\lambda)$ if $\lambda =1$ and $\lambda \sim \text{Gamma}(2, 5)$ ?

```{r eval = FALSE}
plot_gamma(shape = 2, rate = 5)
```

--

.pull-left[
```{r echo = FALSE, fig.align='center', fig.width=5, fig.height=4}
library(bayesrules)
library(ggplot2)
plot_gamma(2,5) +
    geom_point(aes(x = 1, y = dgamma(1, 
                                     shape =2, 
                                     rate =5)), colour="blue") +
  geom_text(aes(label= "(1, ?)", x = 1.25, y = 0.25), 
            color = "blue")
```

]



.pull-right[
```{r}
dgamma(x = 1, shape = 2, rate = 5)
```


]

---

## Tuning Gamma Prior

_Before_ we collected any data, our guess was that the rate of fraud risk calls was most likely around 5 calls per day, but could also reasonably range between 2 and 7 calls per day.

--

In other words, $E(\lambda) = 5$ and $\lambda$ most likely is between 2 and 7. You can use `plot_gamma()` function to try out different gamma distributions.

--

$$E(\lambda) = \frac{s}{r} \approx 5 \; .$$
---

## Tuning Gamma Prior


```{r fig.width = 4, fig.height = 3, message = FALSE, fig.align='center'}
plot_gamma(10,2)
```

---

## The Posterior Model

$$f(\lambda|\vec{y}) \propto f(\lambda)L(\lambda|\vec{y}) = \frac{r^s}{\Gamma(s)} \lambda^{s-1} e^{-r\lambda} \cdot \frac{\lambda^{\sum y_i}e^{-n\lambda}}{\prod y_i!} \;\;\; \text{ for } \lambda > 0.$$
--

$$\begin{split} 
f(\lambda|\vec{y}) 
& \propto \lambda^{s-1} e^{-r\lambda} \cdot \lambda^{\sum y_i}e^{-n\lambda} \\
& = \lambda^{s + \sum  y_i - 1} e^{-(r+n)\lambda} \\
\end{split}$$

--

$$ \lambda|\vec{y} \;  \sim \; \text{Gamma}\bigg(s + \sum  y_i, r + n \bigg) \; .$$


---

__The Gamma-Poisson model__

Let $\lambda > 0$ be an unknown _rate_ parameter and $(Y_1,Y_2,\ldots,Y_n)$ be an independent $\text{Pois}(\lambda)$ sample.
The Gamma-Poisson Bayesian model complements the Poisson structure of data $Y$ with a Gamma prior on $\lambda$:

$$\begin{split}
Y_i | \lambda & \stackrel{ind}{\sim} \text{Pois}(\lambda) \\
\lambda & \sim \text{Gamma}(s, r) \\
\end{split}$$

Upon observing data $\vec{y} = (y_1,y_2,\ldots,y_n)$, the posterior model of $\lambda$ is also a Gamma with updated parameters:

\begin{equation}
\lambda|\vec{y} \; \sim \; \text{Gamma}(s + \sum y_i, \; r + n) \; .
\end{equation}

---

__The Gamma-Poisson model__

```{r fig.height=5}
plot_gamma_poisson(shape = 10, rate = 2, sum_y = 11, n = 4)
```

