---
title: "Sequentiality in Bayesian Analyses"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"
---

class: title-slide

```{r child = "../setup.Rmd"}
```

```{r echo=FALSE, message=FALSE}
library(bayesrules)
library(tidyverse)
```

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
Examples from [bayesrulesbook.com](https://bayesrulesbook.com)

]

---

## Sequential Analysis 

In a sequential Bayesian analysis, a posterior model is updated incrementally as more data comes in.  With the introduction of each new piece of data, the previous posterior model reflecting our understanding prior to observing this data becomes the new prior model.

---

## Let's time travel to the end of 1970

$\pi \sim Beta(14,1)$

--

```{r}
bechdel %>% 
  filter(year == 1970) 
```

---

## The Posterior

```{r}
summarize_beta_binomial(14, 1, y = 1, n = 1)
```

---

### At the end of 1971

$\pi \sim Beta(15,1)$

--

```{r}
bechdel %>% 
  filter(year == 1971) 
```

---

## The Posterior

```{r}
summarize_beta_binomial(15, 1, y = 0, n = 5)
```

---

### At the end of 1972

$\pi \sim Beta(15,6)$

--

```{r}
bechdel %>% 
  filter(year == 1972) 
```

---

## The Posterior

```{r}
summarize_beta_binomial(15, 6, y = 1, n = 3)
```

---

class: middle 

<div align="center">

| Time                | Data         | Model       |
|---------------------|--------------|-------------|
| before the analysis | NA           | Beta(14,1)  |
| at the end of 1970  | Y = 1, n = 1 | Beta(15,1)  |
| at the end of 1971  | Y = 0, n = 5 | Beta(15, 6) |
| at the end of 1972  | Y = 1, n = 3 | Beta(16,8)  |


---

class: middle 

<div align="center">

## Data Order Invariance

| Time                | Data         | Model      |
|---------------------|--------------|------------|
| before the analysis | NA           | Beta(14,1) |
| 1972                | Y = 1, n = 3 | Beta(15,3) |
| 1971                | Y = 0, n = 5 | Beta(15,8) |
| 1970                | Y = 1, n = 1 | Beta(16,8) |


---

class: middle 

<div align="center">

## What if we observed all the data at once?

| Time                | Data         | Model      |
|---------------------|--------------|------------|
| before the analysis | NA           | Beta(14,1) |
| 1970  | Y = 1, n = 1 |            |
|1971  | Y = 0, n = 5 |            |
|1972  | Y = 1, n = 3 |            |
| Total               | Y = 2, n = 9 |            |

---

```{r}
summarize_beta_binomial(14, 1, y = 2, n = 9)
```

---

Let $\theta$ be any parameter of interest with prior pdf $f(\theta)$.  Then a __sequential analysis__ in which we _first_ observe a data point $y_1$ and _then_ a second data point $y_2$ will produce the same posterior model of $\theta$ as if we _first_ observe $y_2$ and *then* $y_1$: 

$$f(\theta | y_1,y_2) = f(\theta|y_2,y_1)\;.$$

Similarly, the posterior model is invariant to whether we observe the data _all at once_ or _sequentially_.


